The Impact of Alternation

### 2017-05-11 draft.tex
ok

Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
\eqiff^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
We omit mentioning the interpretation $\I$ when it is clear from the
context.

@@

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a set of time-stamped variables. For an input event $a \in \Sigma$ and
a formula $\phi$, we write $\Delta^k(\phi,a)$ for the formula obtained
from $\phi$ by simultaneously replacing each state $q \in
\fv{\Bool}{\phi}$ by the formula
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
and, for all $k > 0$, we have $\phi_k \equiv
\Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words accepted by $\A$.

@@

\begin{lemma}\label{lemma:closure}
  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
  L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \eqiff \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \eqiff \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \eqiff \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \eqiff \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \eqiff & \text{ by induction hypothesis} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \eqiff & \text{ by the definition of $\overline{\phi}$} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

@@

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
A formula $\phi(Q,\vec{x})$ \emph{defines} $\Inv$ iff for all $\beta:
Q \rightarrow \booli$ and $\nu : \vec{x} \rightarrow \Data^\I$,
$\I,\beta\cup\nu \models \phi$ if and only if $\nu \in
\Inv(\beta)$. If $\Inv$ satisfies only the first point above, we call
it an \emph{invariant}. Intuitively, a safety invariant maps every
boolean valuation into a set of data valuations, that contains the
initial configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose
data variables are unconstrained, over-approximates the set of
reachable valuations (\ref{it1:inv}) and excludes the valuations
satisfying the acceptance condition (\ref{it2:inv}).

@@

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can now occur negated within the formulae $\phi \in \preds$. We fix
$\preds$ from now on and refer to these formulae as
\emph{predicates}. Given a formula $\phi$, we denote by $\abs{\phi}
\equiv \bigwedge \set{\pi \in \preds \mid \phi \models \pi}$ the
abstraction of $\phi$ w.r.t. the predicates in $\preds$. The abstract
version of the post-image and acceptance condition are defined as
follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \eqiff
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

@@

  (\ref{it1:progress}) Assume that $\pi(u)$ is satisfiable and let
  $\beta_0,\ldots,\beta_n$ and $\nu_0,\ldots,\nu_n$ be boolean and
  data valuations for $Q_0,\ldots,Q_n$ and
  $\vec{x}_0,\ldots,\vec{x}_n$, respectively, such that
  $\I,\bigcup_{i=0}^n \beta_i \cup \bigcup_{i=0}^n \nu_i \models
  \pi(u)$. We build a model for $\Accept{\A}(u) \equiv
  \Post{\A}(\iota,u) \wedge \theta_{n+1}[Q/Q_n]$. Let
  $\phi_0,\phi_1,\ldots,\phi_n$ be the run of $\A$ over $u$. We show
  that $\phi_n(Q,\vec{x}_0\cup\ldots\cup\vec{x}_n)$ has a model, by
  induction on $n\geq0$. For the base case $n=0$, $\phi_0 \equiv \iota
  \equiv \theta_0[Q/Q_0]$ has the model $\beta(q) = \beta_0(q_0)$, for
  all $q \in Q$. For the induction step $n > 0$, we assume that
  $\phi_{n-1}(Q,\vec{x}_0\cup\ldots\vec{x}_{n-1})$ has the model
  $\beta \cup \nu_0 \cup \ldots \nu_{n-1}$, where $\beta(q) =
  \beta_{n-1}(q_{n-1})$, for all $q \in Q$. By the definition of
  $\pi(u)$: \[\theta_n \equiv \bigwedge_{q_{n-1} \in
    \fv{\Bool}{\theta_{n-1}} \cap Q_{n-1}} \left(q_{n-1} \rightarrow
  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]\right)\]

  But since $\I,\beta_{n-1} \cup \beta_n, \nu_{n-1} \cup \nu_n \models
  \theta_n(Q_{n-1}\cup Q_n, \vec{x}_{n-1} \cup \vec{x}_n)$, we have
  that $\I,\beta_n,\nu_{n-1} \cup \nu_n \models
  \Delta(q,a_n)[Q_n/Q,\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  for every $q \in Q$ such that $\beta_{n-1}(q_{n-1}) = \true$. Then
  we have $\I,\beta\cup\bigcup_{i=0}^n\nu_i \models \phi_n$
  where $\beta(q) = \beta_n(q_n)$, for all $q \in Q$ and $\phi_n$ is
  obtained from $\phi_{n-1}$ by replacing each boolean variable $q$ by
  $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$. Observe
  that, because each boolean variable occurs positively in
  $\phi_{n-1}$, the truth value of
  $\Delta(a,q)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$
  under the valuation $\beta\cup\nu_{n-1}\cup\nu_n$ is not important for
  any $q \in Q$ such that $\beta(q) = \false$. We have obtained a
  model for $\phi_n$ and thus for $\Post{\A}(u)$. Since moreover,
  $\theta_{n+1}[Q/Q_n] = \true$ in this model, we obtain a model for
  $\Accept{\A}(u)$. Thus $\pi(u)$ is satisfiable implies that
  $\Accept{\A}(u)$ is satisfiable, which proves the statement of point
  (\ref{it1:progress}).

  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
    interpolant for $\pi(u)$, the following entailments hold: \begin{compactitem}
    \item $\theta_0 \models I_1[Q_0/Q]$, 
      %
    \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
      %
    \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
    \end{compactitem}
    We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$
    by induction on $n \geq 0$. This is sufficient to conclude because
    $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
    a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
    \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we
    have $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv
    \theta_0[Q/Q_0] \models I_1$. For the induction step $n>0$, we
    compute:
    \[\begin{array}{l}
    \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by definition of $\AbsPost{\A}$)} \\
    \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ ($\Asterisk$)} \\
    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
    \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] 
    \end{array}\]
   Let us prove the final point ($\Asterisk$). If
   $\I,\beta_n\cup\nu_{n-1}\cup\nu_n \models
   \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_{n}/Q]$,
   for some valuations $\beta_{n} : Q_{n-1} \rightarrow \booli$ and
   $\nu_i : \vec{x}_i \rightarrow \Data^\I$, for $i=n-1,n$, then we
   build a valuation $\beta_{n-1}$ such that
   $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$,
   as follows. For each occurrence of a subformula $\Delta(q,a_n)$ in
   $\AbsPost{\A}(\iota,a_1\ldots a_{n-1})$ we set
   $\beta_{n-1}(q_{n-1}) = \true$ if $\I,\beta_n\cup\nu_{n-1}\cup\nu_n
   \models \Delta(q,a_n)[Q_n/Q]$ and $\beta_{n-1}(q_{n-1}) = \false$,
   otherwise. Since there are no negated occurrences of such
   subformulae, the definition of $\beta_{n-1}$ is consistent, and the
   check $\I,\beta_{n-1}\cup\beta_n \cup \nu_{n-1}\cup\nu_n \models
   \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n$
   is straightforward. \qed}

We have now all the ingredients to describe the first emptiness
checking semi-algorithm for alternating data
automata. Algorithm\footnote{Though termination is not guaranteed,
  we call it algorithm for conciseness.}  \ref{alg:predabs} builds
an \emph{abstract reachability tree} (ART) whose edges are labeled
with input events and each node $n$ is labeled with the formula
$\AbsPost{\A}(\iota,u)$, where $u \in \Sigma^*$ is the event trace
labeling the path from the root of the tree to $n$. It uses a covering
relation $\lhd$ between nodes in order to ensure that the set of
formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample $u$ is eliminated by computing the
interpolants corresponding to the least unfeasible suffix $v$ of $u$
and recomputing the subtree of the ART rooted in the initial node of
$v$.

@@

  \State $\Call{Close}{n}$
  \label{ln:close-worklist}

@@

  \If{$R(n) \cap F \neq \emptyset$} 
  \label{ln:impact-cex}
  \Comment{counterexample candidate}

@@

  \State let $\tuple{\top,I_1,\ldots,I_k \equiv\bot}$ be an interpolant for $\mu(n)$
  \label{ln:refine-begin}

@@

  \For{$i=1,\ldots,k$}

@@

>>>>

Given formulae $\varphi$ and $\psi$, we say that \emph{$\phi$ entails
  $\psi$}, denoted $\phi \models^\I \psi$ iff $\I,\nu \models \varphi$
implies $\I,\nu \models \psi$, for each valuation $\nu$, and $\phi
\iffeq^\I \psi$ iff $\phi \models^\I \psi$ and $\psi \models^\I \phi$.
We omit mentioning the interpretation $\I$ when it is clear from the
context.

@@

Formally, let $\vec{x}_k = \set{x_k \mid x \in \vec{x}}$, $k\geq0$, be
a set of time-stamped variables. For an input event $a \in \Sigma$ and
a formula $\phi$, we write $\Delta(\phi)$ and, respectively,
$\Delta^k(\phi,a)$, for the formula obtained from $\phi$ by
simultaneously replacing each state $q \in \fv{\Bool}{\phi}$ by the
formula $\Delta(q,a)$, respectively
$\Delta(q,a)[\vec{x}_k/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}]$, for
all $k\geq0$. Given a word $w = (a_1,\nu_1)(a_2,\nu_2) \ldots
(a_n,\nu_n)$, the \emph{run} of $\A$ over $w$ is the sequence of
formulae $\phi_0,\phi_1,\ldots,\phi_n$, where $\phi_0 \equiv \iota$
and, for all $k > 0$, we have $\phi_k \equiv
\Delta^k(\phi_{k-1},a_k)$. Next, we slightly abuse notation and write
$\Delta(\iota,a_1,\ldots,a_n)$ for the formula
$\phi_n(\vec{x}_0,\ldots,\vec{x}_n)$ above. We say that $\A$
\emph{accepts} $w$ iff $\I,\nu \models \Delta(\iota,a_1,\ldots,a_n)$,
where $\nu$ is some valuation that maps:\begin{inparaenum}[(1)]
%
\item each $x \in \vec{x}_k$ to $\nu_k(x)$, for all $k\in[1,n]$, 
%
\item each $q \in \fv{\Bool}{\phi_n} \cap F$ to $\top$ and 
%
\item each $q \in \fv{\Bool}{\phi_n} \setminus F$ to $\bot$.
\end{inparaenum}
The language of $\A$ is the set $L(\A)$ of words accepted by $\A$.

@@

\begin{lemma}\label{lemma:closure}
  Given automata $\A_i = \tuple{\vec{x},Q_i,\iota_i,F_i,\Delta_i}$,
  for $i=1,2$, such that $Q_1 \cap Q_2 = \emptyset$, we have
  $L(\A_\cup) = L(\A_1) \cup L(\A_2)$, $L(\A_\cap) = L(\A_1) \cap
  L(\A_2)$ and $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus
  L(\A_1)$.
\end{lemma}
\proof{ We prove $L(\A_\cup) = L(\A_1) \cup L(\A_2)$ first, the proof
  for $\A_\cap$ being analogous. Let $w = (a_1,\nu_1) \ldots
  (a_n,\nu_n)$ be a word, where $n=0$ corresponds to the empty
  word. We prove by induction on $n\geq0$ that $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_n) \iffeq \Delta(\iota_1,a_1\ldots a_n) \vee
  \Delta(\iota_2,a_1\ldots a_n)$. The case $n=0$ follows from the
  definition of the initial configuration of $\A_\cup$. For the
  inductive step $n>0$, $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_n)$
  is obtained from $\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1})$ by
  replacing each variable $q \in \fv{\Bool}{\Delta(\iota_1 \vee
    \iota_2,a_1\ldots a_{n-1})}$ with
  $\Delta(q,a_n)[\vec{x}_{n-1}/\overline{\vec{x}},\vec{x}_n/\vec{x}]$,
  denoted $\Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots
  a_{n-1}),a_n)$. Since by induction hypothesis, $\Delta(\iota_1 \vee
  \iota_2,a_1\ldots a_{n-1}) \iffeq \Delta(\iota_1,a_1\ldots a_{n-1})
  \vee \Delta(\iota_2,a_1\ldots a_{n-1})$, we obtain:
  \[\begin{array}{ll}
  \Delta^n(\Delta(\iota_1 \vee \iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta^n(\Delta(\iota_1,a_1\ldots a_{n-1}),a_n) \vee \Delta^n(\Delta(\iota_2,a_1\ldots a_{n-1}),a_n) & \iffeq \\
  \Delta(\iota_1,a_1\ldots a_{n}) \vee \Delta(\iota_2,a_1\ldots a_{n}) \enspace.
  \end{array}\] 
  To prove $L(\overline{\A_1}) = \Sigma[\vec{x}]^* \setminus L(\A_1)$,
  let $w = (a_1,\nu_1) \ldots (a_n,\nu_n)$ be a word and show that
  $\overline{\Delta}(\iota_1,a_1 \ldots a_n) =
  \overline{\Delta(\iota_1,a_1 \ldots a_n)}$ by induction on
  $n\geq0$. The case $n=0$ is immediate, because $\fv{}{\iota_1}
  \subseteq Q$ and thus $\overline{\iota_1} = \iota_1$. For the case
  $n>0$, we compute:
  \[\begin{array}{lcl}
  \overline{\Delta}^n(\overline{\Delta}(\iota_1,a_1\ldots a_{n-1}),a_n) & \iffeq & \text{ by induction hypothesis} \\
  \overline{\Delta}^n(\overline{\Delta(\iota_1,a_1\ldots a_{n-1})},a_n) & \iffeq & \text{ by the definition of $\overline{\phi}$} \\
  \overline{\Delta(\iota_1,a_1\ldots,a_n)}
  \end{array}\]
  Let $\nu,\nu' : (Q \cup \bigcup_{i=0}^n\vec{x}_i) \rightarrow (\booli
  \cup \Data^\I)$ be valuations such that: \begin{compactitem}
  \item $\nu(q) = \top$ and $\nu'(q) = \bot$, for each $q \in F$, 
    %
  \item $\nu(q) = \bot$ and $\nu'(q) = \top$, for each $q \in Q \setminus F$, 
    %
  \item $\nu(x) = \nu'(x)$, for each $x \in \vec{x}_0$, 
    %
  \item $\nu(x) = \nu'(x) = \nu_i(x)$, for each $x \in \vec{x}_i$ and each $i
    \in [1,n]$.
  \end{compactitem}
  By Proposition \ref{prop:overline}, we have $\I,\nu \models
  \Delta(\iota_1,a_1 \ldots a_n) \iff \I,\nu' \not\models
  \overline{\Delta(\iota_1,a_1 \ldots a_n)} \iff \I,\nu' \not\models
  \overline{\Delta}(\iota_1,a_1 \ldots a_n)$. Thus $w \in L(\A_1) \iff
  w \not\in L(\overline{\A_1})$, i.e. $L(\overline{\A_1}) =
  \Sigma[\vec{x}]^* \setminus L(\A_1)$. \qed}

@@

A \emph{safety invariant} for $\A$ is a function $\Inv : (Q \mapsto
\booli) \rightarrow 2^{\vec{x} \mapsto \Data^\I}$ such that, for every
boolean valuation $\beta : Q \rightarrow \booli$, every valuation $\nu
: \vec{x} \mapsto \Data^\I$ of the data variables and every finite
sequence $u \in \Sigma^*$ of input events, the following
hold: \begin{compactenum}
%% \item $\beta \models \iota \Rightarrow \nu_0 \in \Inv(\beta)$, where
%%   $\nu_0(x) = \mathbf{0}$, for all $x \in \vec{x}$.
%
\item\label{it1:inv} $\I,\beta \cup \nu \models \Post{\A}(\iota,u)
  \Rightarrow \nu \in \Inv(\beta)$, and
%
\item\label{it2:inv} $\nu \in \Inv(\beta) \Rightarrow \I,\beta \cup
  \nu \not\models \Accept{\A}(u)$.
\end{compactenum}
If $\Inv$ satisfies only the first point above, we call it an
\emph{invariant}. Intuitively, a safety invariant maps every boolean
valuation into a set of data valuations, that contains the initial
configuration $\iota = \Post{\A}(\iota,\varepsilon)$, whose data
variables are unconstrained, over-approximates the set of reachable
valuations (\ref{it1:inv}) and excludes the valuations satisfying the
acceptance condition (\ref{it2:inv}).  A formula $\phi(Q,\vec{x})$
\emph{defines} $\Inv$ iff for all $\beta: Q \rightarrow \booli$ and
$\nu : \vec{x} \rightarrow \Data^\I$, we have $\I,\beta\cup\nu \models
\phi$ iff $\nu \in \Inv(\beta)$.

@@

Formally, let $\preds$ be a finite set of formulae with variables from
$Q \cup \vec{x}$, such that $\bot \in \preds$ and the variables in $Q$
can now occur negated within the formulae $\phi \in \preds$. We fix
$\preds$ from now on and refer to these formulae as
\emph{predicates}. Given a formula $\phi$, we denote by $\abs{\phi}
\equiv \bigwedge \set{\pi \in \preds \mid \phi \models \pi}$ the
abstraction of $\phi$ w.r.t. the predicates in $\preds$. The abstract
version of the post-image and acceptance condition are defined as
follows, for any $u \in \Sigma^*$:
\[\begin{array}{l}
\AbsPost{\A}(\phi,\varepsilon) \equiv \phi \hspace*{5mm}
\AbsPost{\A}(\phi,ua) \equiv \abs{(\Post{\A}(\AbsPost{\A}(\phi,u),a))} \\
\AbsAccept{\A}(u) \equiv \AbsPost{\A}(\iota,u) \wedge \bigwedge_{q \in Q \setminus F} 
(q \rightarrow \bot)
\end{array}\]
\begin{lemma}\label{lemma:abstract-invariant}
  For any bijection $\mu : \nat \rightarrow \Sigma^*$, there exists
  $k>0$ such that $\bigvee_{i=1}^k \AbsPost{\A}(\iota,\mu(i))$ defines
  an invariant for $\A$. 
\end{lemma}
\proof{ It is sufficient to show that there exists $k\geq0$ such that for
  all $u \in \Sigma^*$ there exists $i \in [0,k]$ such that
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,\mu(i))$. We have
  $\Post{\A}(\iota,u) \models \AbsPost{\A}(\iota,u)$ for all $u \in
  \Sigma^*$. But since $\preds$ is a finite set, also the set
  $\{\AbsPost{\A}(\iota,u) \mid u \in \Sigma^*\}$ is finite. Thus
  there exists $k \geq 0$ such that, for all $u \in \Sigma^*$ there
  exists $i \in [0,k]$ such that $\AbsPost{\A}(\iota,u) \iffeq
  \AbsPost{\A}(\iota,\mu(i))$, which concludes the proof. \qed}

@@

\begin{proposition}\label{prop:asterisk}
  Given a formula $\phi \in \Form^+(Q,\vec{x})$ and $a \in \Sigma$, we
  have \(\Delta(\phi,a) \iffeq \exists Q' ~.~ \phi[Q'/Q] \wedge
  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))\).
\end{proposition}
\proof{``$\Rightarrow$'' If $\I,\beta\cup\overline\nu\cup\nu \models
  \Delta^n(\phi,a)$, for some valuations $\beta : Q \rightarrow
  \booli$ and $\overline\nu : \overline{\vec{x}} \rightarrow
  \Data^\I$, $\nu : \vec{x} \rightarrow \Data^\I$, then we build a
  valuation $\beta' : Q' \rightarrow \booli$ such that
  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$. For each
  occurrence of a formula $\Delta(q,a)$ in $\Delta(\phi,a)$ we set
  $\beta'(q') = \true$ if $\I,\beta\cup\overline\nu\cup\nu \models
  \Delta(q,a)$ and $\beta'(q') = \false$, otherwise. Since there are
  no negated occurrences of such subformulae, the definition of
  $\beta'$ is consistent, and the check
  $\I,\beta'\cup\beta\cup\overline\nu\cup\nu \models \phi[Q'/Q] \wedge
  \bigwedge_{q \in Q} (q' \rightarrow \Delta(q,a))$ is
  immediate. $''\Leftarrow''$ This direction is an easy check. \qed}

@@

  (\ref{it1:progress}) We apply Proposition \ref{prop:asterisk}
  recursively and get: \[\Post{\A}(\iota,u)[Q_n/Q,\vec{x}_n/\vec{x}]
  \iff \exists Q_0 \ldots \exists Q_{n-1} \exists \vec{x}_0 \ldots
  \exists \vec{x}_{n-1} ~.~ \bigwedge_{i=0}^n \theta_i\] Assuming that
  $\pi(u) \equiv \bigwedge_{i=0}^{n+1} \theta_i$ is satisfiable, we
  obtain a model for $\Accept{\A}(u) \equiv \Post{\A}(\iota,u) \wedge
  \theta_{n+1}[Q/Q_n]$.

  (\ref{it2:progress}) If $\tuple{\top,I_1,\dots,I_{n+1},\bot}$ is an
  interpolant for $\pi(u)$, the following entailments
  hold: \begin{compactitem}
  \item $\theta_0 \models I_1[Q_0/Q]$,
    %
  \item $I_k[Q_{k-1}/Q,\vec{x}_{k-1}/\vec{x}] \wedge \theta_k \models I_{k+1}[Q_k/Q,\vec{x}_k/\vec{x}]$, $\forall k\in[1,n]$. 
    %
  \item $I_{n+1}[Q_{n+1}/Q] \wedge \theta_{n+1} \models \bot$. 
  \end{compactitem}
  We prove that $\AbsPost{\A}(\iota,a_1\ldots a_n) \models I_{n+1}$ by
  induction on $n \geq 0$. This is sufficient to conclude because
  $\AbsAccept{\A}(a_1\ldots a_n) \equiv \AbsPost{\A}(\iota,a_1\ldots
  a_n) \wedge \theta_{n+1}[Q/Q_n] \models I_{n+1} \wedge
  \theta_{n+1}[Q/Q_n] \models \bot$. For the base case $n=0$, we have
  $\AbsPost{\A}(\iota,\varepsilon) \equiv \iota \equiv \theta_0[Q/Q_0]
  \models I_1$. For the induction step $n>0$, we compute:
  \[\begin{array}{l}
  \AbsPost{\A}(\iota,a_1\ldots a_n)[Q_n/Q] \equiv \text{ (by definition of $\AbsPost{\A}$)} \\
  \exists \vec{x}_{n-1} ~.~ \abs{\Delta^n(\AbsPost{\A}(\iota,a_1\ldots a_{n-1}),a_n)}[Q_n/Q] \models \text{ (by Prop. \ref{prop:asterisk})} \\
  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ \AbsPost{\A}(\iota,a_1\ldots a_{n-1})[Q_{n-1}/Q] \wedge \theta_n \models \text{ (ind. hyp.) } \\
  \exists Q_{n-1}  \exists \vec{x}_{n-1} ~.~ I_{n-1}[Q_{n-1}/Q] \wedge \theta_n \models I_n[Q_n/Q] \hfill\text{\qed}
  \end{array}\]}
We have now all the ingredients to describe the first emptiness
checking semi-algorithm for alternating data
automata. Algorithm\footnote{Though termination is not guaranteed, we
  call it algorithm for conciseness.}  \ref{alg:predabs} builds an
\emph{abstract reachability tree} (ART) whose edges are labeled with
input events and each node $n$ is labeled with the formula
$\AbsPost{\A}(\iota,u)$, where $u \in \Sigma^*$ is the event trace
labeling the path from the root of the tree to $n$. It uses a covering
relation $\lhd$ between nodes in order to ensure that the set of
formulae labeling the nodes in the ART forms an antichain. Any
spurious counterexample $u$ is eliminated by computing the
interpolants corresponding to the least unfeasible suffix $v$ of $u$
and recomputing the subtree of the ART rooted in the initial node of
$v$.

@@

@@

@@

  \State let $\tuple{\top,I_0,\ldots,I_k,\bot}$ be an
  interpolant for $\mu(n)$ with $I_0,\ldots,I_k \in
  \Form^+(Q,\vec{x})$
  \label{ln:refine-begin}

@@

  \For{$i=0,\ldots,k$}

@@

  \EndIf % $\neg b$

  \EndIf % $\Lambda(n_i) \not\models I_i$
